A note on the instability and degeneracy of deep learning models

### 2017-09-30 writing/paper-note/chapters/instab.Rmd
what a mess

To define a measure of instability in FOES models, it is useful to consider the behavior of  data  models $P_{\theta_N}$, again supported on a set $\mathcal{X}^N$ of outcomes for $\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)$,  in connection to the sample size $N$.  A relevant quantity to this end is a log-ratio of extremal probabilities (LREP), defined as
\begin{align}
(\#eq:elpr)
 \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
\end{align}
based on maximum and minimal model probabilities. In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation for $N$ random variables where the log-ratio (\ref{eq:elpr}) is overly large relative to the sample size $N$. That is, a sequence of FOES probability models $P_{\boldsymbol \theta_N}$ results in specifying the distribution of observations $\boldsymbol X=(X_1,\ldots,X_N)$ for each sample size $N \geq 1$ and instability will generally occur among these models whenever the corresponding LREP \@ref(eq:elpr) grows faster than $N$. This leads to the following definition.

```{definition, name="S-unstable FOES model", label="instabFSFS", echo=TRUE}
A FOES model formulation for $\boldsymbol X_N=(X_1,\ldots,X_N)$ is *Schweinberger-unstable* or *S-unstable* if
\begin{equation}
(\#eq:Sun)
\lim \limits_{N \rightarrow \infty} \frac{1}{N} \REP(\boldsymbol \theta_N) \equiv \lim \limits_{N \rightarrow \infty} \frac{1}{N}\log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right] = \infty
\end{equation}
 as the number of variables increases ($N \rightarrow \infty$).
```
In other words, a model is S-unstable if $\REP(\boldsymbol \theta_N)/N$ is an unbounded sequence of sample size  $N$; namely, given any $C > 0$, there exists an integer $N_C > 0$ so that $\REP(\boldsymbol \theta_N)/N > C$ holds for all $N \ge N_C$. A FOES model formulation may be termed S-stable if it fails to be S-unstable, i.e., if $\sup_{N \geq 1}\REP(\boldsymbol \theta_N)/N $ is bounded.

This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing possibility non-exponential family models (e.g., RBM and DBM models in Sections \@ref(restricted-boltzmann-machines) - \@ref(deep-learning) as well as a potentially increasing number $d(N)$ of parameters through the parameter sequence $\boldsymbol \theta_N\in \mathbb{R}^{d(N)}$. While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf. Section \@ref(discrete-exponential-family-models) considered there. Section \ref@(illustrations) provides several examples  of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or parameterization. We next describe several potentially undesirable features associated with unstable models.

```{remark, echo=TRUE}
In the definition \@ref(eq:Sun) of S-instability, we note that the numerical measure $\REP(\boldsymbol \theta_N)/N$ of model instability is invariant to *independent  replications* of data. That is, let $M \geq 1$ denote a possible number of replications and consider data $\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)$ formed by $\{ \boldsymbol X^{(j)}_N\}_{j=1}^M$ as $M$  iid replications of a random vector $\boldsymbol X_N=(X_1,\ldots,X_N)$, where the latter follows a FOES model with probabilities \(P_{ \boldsymbol \theta_N}(\boldsymbol x)>0\), $\boldsymbol x\in\mathcal{X}^N$. This leads to a joint model, say $P_{ \boldsymbol \theta_N}(\boldsymbol y)$, $\boldsymbol y\in\mathcal{X}^{NM}$, for $\boldsymbol Y_{N,M}$ consisting of $N*M$ random variables in total. Then, the LREP for $\boldsymbol Y_{N,M}$, scaled by associated size, is given by
\begin{align*}
\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}( \boldsymbol\theta_N ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)} \right] \\
&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
\end{align*}
where $\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N )  \equiv \REP( \boldsymbol\theta_N )$ denotes the log-ratio of extremal probabilities for $\boldsymbol X_N$ defined from \@ref(eq:elpr). That is, due to iid properties, the sample-size corrected  LREP for $\boldsymbol Y_{N,M}$ equals the analog, $\REP(\boldsymbol\theta_N )/N$, from the underlying common data model for $\boldsymbol X_N$ alone, regardless of the level $M \geq 1$ of independent replication. Consequently, the definition of an S-unstable  model  is  unaffected by independent replication and all instability properties  may be characterized by those of one observation  from the common FOES model. For computational purposes, this aspect also implies that if the original data $\boldsymbol X_N=(X_1,\ldots,X_N)$ in a FOES model consist of $N$ iid random variables, then the size-scaled log-ratio \@ref(eq:elpr)  may be calculated as
$$
\frac{1}{N}\REP( \boldsymbol\theta_N ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} \right]
$$
based on the extremal probabilities of just one random variable $X_1$.
```

@@

```{theorem, label="instab-elpr", echo=TRUE}
Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models.

(i) For any integer $N \geq 1$ and any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}), then
    $$ \DN(\boldsymbol \theta_N) > C,$$
    or probabilities from a one component change in some outcome have log-ratio exceeding $C$.
(ii) Suppose the FOES model sequence is S-unstable. Then, for all large $N$ and given any arbitrary $C>0$, there exist outcomes $\boldsymbol x,\boldsymbol x^*\in\mathcal{X}^N$, differing by one component, such that
    $$ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} > \exp[N C]. $$
```

Theorem \@ref(instab-elpr)(i) is a  non-asymptotic result, which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities \@ref(eq:elpr) being too large relative to the associated sample size $N$. If so, Theorem \ref@(instab-elpr)(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for very small differences among some data configurations, a property that intuitively captures a notion of instability.  Furthermore, and perhaps more seriously under Theorem \@ref(instab-elpr)(ii),
S-unstable models can never have universally bounded changes in probability among single component variations in data configurations. While not all one component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property.  As a consequence, unstable probability structure may exhibit extreme sensitivity through large peaks and troughs over the sample space.

Additionally, S-unstable FOES model sequences are also connected to degenerate models, where {\it degeneracy} involves assigning essentially all probability to modes within the sample space, which could potentially represent a  small subset among the totality of outcomes. For perspective, note that differing sizes of the scaled log-ratio $\REP(\boldsymbol \theta_N)/N$ from \@ref(eq:elpr) induce a spectrum of levels of instability/stability and Theorem \@ref(instab-elpr) indicates increasing sensitivity of model probabilities as \@ref(eq:elpr) increases. Furthermore, as the instability measure grows and the log-ratio $\REP(\boldsymbol \theta_N)/N$ diverges, as in the definition \@ref(eq:Sun) of S-unstable models, then a FOES model sequence  will become degenerate. Theorem \@ref(degenFOES) provides a formal statement of such degeneracy due to instability. For a given $0 < \epsilon < 1$, define a $\epsilon$-modal set of   outcomes as
\begin{equation}
(\#eq:mode)
\mathcal{M}_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{\boldsymbol \theta_N}(\boldsymbol y) + \epsilon\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{\boldsymbol \theta_N}(\boldsymbol y) \right\}.
\end{equation}

@@

```{mcpsr, name="Reciprocal Probabilities from Parameter Sign Reversal", echo=TRUE}
Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, represent a sequence of FOES models. For each $N \geq 1$ and any outcome $\boldsymbol x \in \mathcal{X}^N$, suppose it holds that
$$
P_{\boldsymbol \theta_N}(\boldsymbol x)  \cdot P_{-\boldsymbol \theta_N}(\boldsymbol x) \;=\;   \max\limits_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)\cdot \min\limits_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y),
$$
where $\max_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)$ and $\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)$ denote the maximum and minimum probabilities under parameters $\boldsymbol \theta_N$   and $-\boldsymbol \theta_N$, respectively.
```

The above model condition incorporates many standard parameterizations and follows, for instance, whenever $P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y)  = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}$ holds for outcomes $\boldsymbol x, \boldsymbol y \in\mathcal{X}^N$ in a FOES model. For instance, this latter condition is fulfilled for all linear exponential families from Section \@ref(discrete-exponential-family-models) (e.g., \@ref(eq:mod1)-\@ref(eq:mod2)) as well as all network models from Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning) (e.g., \@ref(eq:RBM1)-\@ref(eq:RBM2)). When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the following extension of Theorem \@ref(degenFOES}).

```{corollary, label="sign", echo=TRUE}
Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models satisfying model condition PSR. If the models $P_{\boldsymbol \theta_N}$  are additionally S-unstable, then

(i) the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
(ii) and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N} $ under $\boldsymbol \theta_N$ from (\ref{eqn:mode}), with $0<\epsilon<1$, it holds under $- \boldsymbol \theta_N$ that
    $$ P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \mbox{as $N\to \infty$},$$
    while, by Theorem \@ref(degenFOES), $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $ \boldsymbol X_N = (X_1,\ldots,X_N)$ under $  \boldsymbol \theta_N$.
```

For unstable models, Corollary \@ref(sign) shows that shifts in parameters around zero (i.e., from $\boldsymbol \theta_N$ to $-\boldsymbol \theta_N$) can induce extreme changes in probability among subsets of the sample space, as another manifestation  of instability and hyper-sensitivity in probability structure. For one-parameter exponential families, involving a fixed real-valued linear parameter  $\boldsymbol \theta_N = \theta \in \mathbb{R}$ and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$ in \@ref(eq:expo), @schweinberger2011instability [Theorem 3] proved a result similar in spirit, though based on a characterization there in terms of maximum $U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ and minimal $L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ values of the sufficient statistic. For this case in particular,  mode sets have specific, and essentially complementary, forms over positive and negative parameters, namely, $\mathcal{M}_{\epsilon, \boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x)  >  (1-\epsilon) U_N + \epsilon L_N \}$ and $\mathcal{M}_{\epsilon, -\boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x)  <  \epsilon U_N + (1-\epsilon) L_N \}$ for any $\boldsymbol \theta_N>0$, and
 @schweinberger2011instability [Theorem 3] showed each mode set collects all mass, under positive and negative parameters, respectively, with unstable models of this exponential type. However, for all unstable FOES models, Corollary \@ref(sign) generalizes the same principle that unstable models can push all probability to different, and in fact disjoint, parts of the sample space, depending on how parameters fall with respect to zero. This feature can numerically complicate likelihood manipulations, such as maximization or MCMC-based Bayes posterior sampling, as further discussed in Section \@ref(statistical-consequences).

# Implications

For a large class of models that covers a broad range of applications (including "deep learning"), we have developed a formal definition of instability in model probability structure and elucidated multiple consequences of instability. We have shown for FSFS models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which potentially could be small. The FSFS model class is quite broad and, particularly in developing FSFS models for large data sets, some caution should be used in parameter specification to control effects of model instability.

>>>>

To define a measure of instability in FOES models, it is useful to consider the behavior of  data  models $P_{\theta_N}$, again supported on a set $\mathcal{X}^N$ of outcomes for $\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)$,  in connection to the sample size $N$.  A relevant quantity to this end is a log-ratio of extremal probabilities (LREP), defined as
\begin{align}
(\#eq:elpr)
 \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
\end{align}
based on maximum and minimal model probabilities. In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation for $N$ random variables where the log-ratio \@ref(eq:elpr) is overly large relative to the sample size $N$. That is, a sequence of FOES probability models $P_{\boldsymbol \theta_N}$ results in specifying the distribution of observations $\boldsymbol X=(X_1,\ldots,X_N)$ for each sample size $N \geq 1$ and instability will generally occur among these models whenever the corresponding LREP \@ref(eq:elpr) grows faster than $N$. This leads to the following definition.

```{definition, name="S-unstable FOES model", label="instabFSFS", echo=TRUE}
A FOES model formulation for $\boldsymbol X_N=(X_1,\ldots,X_N)$ is *Schweinberger-unstable* or *S-unstable* if
\begin{equation}
(\#eq:Sun)
\lim \limits_{N \rightarrow \infty} \frac{1}{N} \REP(\boldsymbol \theta_N) \equiv \lim \limits_{N \rightarrow \infty} \frac{1}{N}\log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right] = \infty
\end{equation}
 as the number of variables increases ($N \rightarrow \infty$).
```
In other words, a model is S-unstable if $\REP(\boldsymbol \theta_N)/N$ is an unbounded sequence of sample size $N$; namely, given any $C > 0$, there exists an integer $N_C > 0$ so that $\REP(\boldsymbol \theta_N)/N > C$ holds for all $N \ge N_C$. A FOES model formulation may be termed S-stable if it fails to be S-unstable, i.e., if $\sup_{N \geq 1}\REP(\boldsymbol \theta_N)/N$ is bounded.

This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing possibility non-exponential family models (e.g., RBM and DBM models in Sections \@ref(restricted-boltzmann-machines) - \@ref(deep-learning) as well as a potentially increasing number $d(N)$ of parameters through the parameter sequence $\boldsymbol \theta_N\in \mathbb{R}^{d(N)}$. While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf. Section \@ref(discrete-exponential-family-models) considered there. Section \@ref(illustrations) provides several examples  of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or parameterization. We next describe several potentially undesirable features associated with unstable models.

```{remark, label="rmk1", echo=TRUE}
In the definition \@ref(eq:Sun) of S-instability, we note that the numerical measure $\REP(\boldsymbol \theta_N)/N$ of model instability is invariant to *independent  replications* of data. That is, let $M \geq 1$ denote a possible number of replications and consider data $\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)$ formed by $\{ \boldsymbol X^{(j)}_N\}_{j=1}^M$ as $M$  iid replications of a random vector $\boldsymbol X_N=(X_1,\ldots,X_N)$, where the latter follows a FOES model with probabilities \(P_{ \boldsymbol \theta_N}(\boldsymbol x)>0\), $\boldsymbol x\in\mathcal{X}^N$. This leads to a joint model, say $P_{ \boldsymbol \theta_N}(\boldsymbol y)$, $\boldsymbol y\in\mathcal{X}^{NM}$, for $\boldsymbol Y_{N,M}$ consisting of $N*M$ random variables in total. Then, the LREP for $\boldsymbol Y_{N,M}$, scaled by associated size, is given by
\begin{align*}
\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}( \boldsymbol\theta_N ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)} \right] \\
&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
\end{align*}
where $\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N )  \equiv \REP( \boldsymbol\theta_N )$ denotes the log-ratio of extremal probabilities for $\boldsymbol X_N$ defined from \@ref(eq:elpr). That is, due to iid properties, the sample-size corrected  LREP for $\boldsymbol Y_{N,M}$ equals the analog, $\REP(\boldsymbol\theta_N )/N$, from the underlying common data model for $\boldsymbol X_N$ alone, regardless of the level $M \geq 1$ of independent replication. Consequently, the definition of an S-unstable  model  is  unaffected by independent replication and all instability properties  may be characterized by those of one observation  from the common FOES model. For computational purposes, this aspect also implies that if the original data $\boldsymbol X_N=(X_1,\ldots,X_N)$ in a FOES model consist of $N$ iid random variables, then the size-scaled log-ratio \@ref(eq:elpr)  may be calculated as
$$
\frac{1}{N}\REP( \boldsymbol\theta_N ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} \right]
$$
based on the extremal probabilities of just one random variable $X_1$.
```

@@

```{theorem, label="instab-elpr", echo=TRUE}
Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models.
\begin{enumerate}[(i)]
\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}, then
    $$ \DN(\boldsymbol \theta_N) > C,$$
    or probabilities from a one component change in some outcome have log-ratio exceeding $C$.
\item Suppose the FOES model sequence is S-unstable. Then, for all large $N$ and given any arbitrary $C>0$, there exist outcomes $\boldsymbol x,\boldsymbol x^*\in\mathcal{X}^N$, differing by one component, such that
    $$
    \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} > \exp[N C].
    $$
\end{enumerate}
```

Theorem \@ref(thm:instab-elpr)(i) is a non-asymptotic result, which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities \@ref(eq:elpr) being too large relative to the associated sample size $N$. If so, Theorem \@ref(thm:instab-elpr)(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for very small differences among some data configurations, a property that intuitively captures a notion of instability.  Furthermore, and perhaps more seriously under Theorem \@ref(thm:instab-elpr)(ii),
S-unstable models can never have universally bounded changes in probability among single component variations in data configurations. While not all one component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property.  As a consequence, unstable probability structure may exhibit extreme sensitivity through large peaks and troughs over the sample space.

Additionally, S-unstable FOES model sequences are also connected to degenerate models, where *degeneracy* involves assigning essentially all probability to modes within the sample space, which could potentially represent a  small subset among the totality of outcomes. For perspective, note that differing sizes of the scaled log-ratio $\REP(\boldsymbol \theta_N)/N$ from \@ref(eq:elpr) induce a spectrum of levels of instability/stability and Theorem \@ref(thm:instab-elpr) indicates increasing sensitivity of model probabilities as \@ref(eq:elpr) increases. Furthermore, as the instability measure grows and the log-ratio $\REP(\boldsymbol \theta_N)/N$ diverges, as in the definition \@ref(eq:Sun) of S-unstable models, then a FOES model sequence  will become degenerate. Theorem \@ref(thm:degenFOES) provides a formal statement of such degeneracy due to instability. For a given $0 < \epsilon < 1$, define a $\epsilon$-modal set of   outcomes as
\begin{equation}
(\#eq:mode)
\mathcal{M}_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{\boldsymbol \theta_N}(\boldsymbol y) + \epsilon\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{\boldsymbol \theta_N}(\boldsymbol y) \right\}.
\end{equation}

@@

**Model Condition PSR** *(Reciprocal Probabilities from Parameter Sign Reversal)*: Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, represent a sequence of FOES models. For each $N \geq 1$ and any outcome $\boldsymbol x \in \mathcal{X}^N$, suppose it holds that
$$
P_{\boldsymbol \theta_N}(\boldsymbol x)  \cdot P_{-\boldsymbol \theta_N}(\boldsymbol x) \;=\;   \max\limits_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)\cdot \min\limits_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y),
$$
where $\max_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)$ and $\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)$ denote the maximum and minimum probabilities under parameters $\boldsymbol \theta_N$ and $-\boldsymbol \theta_N$, respectively.

The above model condition incorporates many standard parameterizations and follows, for instance, whenever $P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y)  = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}$ holds for outcomes $\boldsymbol x, \boldsymbol y \in\mathcal{X}^N$ in a FOES model. For instance, this latter condition is fulfilled for all linear exponential families from Section \@ref(discrete-exponential-family-models) (e.g., \@ref(eq:mod1)-\@ref(eq:mod2)) as well as all network models from Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning) (e.g., \@ref(eq:RBM1)-\@ref(eq:RBM2)). When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the following extension of Theorem \@ref(thm:degenFOES}).

```{corollary, label="sign", echo=TRUE}
Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models satisfying model condition PSR. If the models $P_{\boldsymbol \theta_N}$  are additionally S-unstable, then
\begin{enumerate}[(i)]
\item the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $- \boldsymbol \theta_N$ that
    $$ 
    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \mbox{as $N\to \infty$},
    $$
    while, by Theorem \ref{thm:degenFOES}, $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\boldsymbol \theta_N$.
\end{enumerate}  
```

For unstable models, Corollary \@ref(cor:sign) shows that shifts in parameters around zero (i.e., from $\boldsymbol \theta_N$ to $-\boldsymbol \theta_N$) can induce extreme changes in probability among subsets of the sample space, as another manifestation  of instability and hyper-sensitivity in probability structure. For one-parameter exponential families, involving a fixed real-valued linear parameter  $\boldsymbol \theta_N = \theta \in \mathbb{R}$ and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$ in \@ref(eq:expo), @schweinberger2011instability [Theorem 3] proved a result similar in spirit, though based on a characterization there in terms of maximum $U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ and minimal $L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ values of the sufficient statistic. For this case in particular,  mode sets have specific, and essentially complementary, forms over positive and negative parameters, namely, $\mathcal{M}_{\epsilon, \boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x)  >  (1-\epsilon) U_N + \epsilon L_N \}$ and $\mathcal{M}_{\epsilon, -\boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x)  <  \epsilon U_N + (1-\epsilon) L_N \}$ for any $\boldsymbol \theta_N>0$, and
 @schweinberger2011instability [Theorem 3] showed each mode set collects all mass, under positive and negative parameters, respectively, with unstable models of this exponential type. However, for all unstable FOES models, Corollary \@ref(cor:sign) generalizes the same principle that unstable models can push all probability to different, and in fact disjoint, parts of the sample space, depending on how parameters fall with respect to zero. This feature can numerically complicate likelihood manipulations, such as maximization or MCMC-based Bayes posterior sampling, as further discussed in Section \@ref(implications).

```{remark, name="need to briefly finish", echo=TRUE}
extensions of sensitivity... basically, model condition PSR and Corollary \ref{cor:sign} can be extended to cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ are not all changed in sign ($-\boldsymbol \theta_N$) but, more generally, are instead altered to another parameter configuration $\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )$ involving a switch in sign only among some model parameters $\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen. If a change sign occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then an unstable model under $\boldsymbol \theta_N$ will then imply that many more unstable models exist  over a broad spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$.
```

# Illustrations

Model instability can depend intricately on how functions of parameters and data $\boldsymbol X_N=(X_1,\ldots,X_n)$ are combined in the formulation of the model probabilities, though some general causes may be identified. As one issue, a broad parameter space (or wide interpretation of this space) may admit some parameters as technically valid that have an undue and often undesirable impact on the model structure for a prescribed data size $N$. In this case, both the size and dimension of model parameters can be problematic and induce instability. In combination to this last point, further causes of instability may also be traced to the magnitude of statistics in the model. Potentially massive, and thereby unstable, statistics were the primary focus of instability studies of @schweinberger2011instability for certain discrete exponential models having parameters/statistics of fixed dimension. However, as shown in the following, bounded statistics may still lead to instability if the parameter dimension is high. We next provides some examples to illustrate instability in FOES models, which also suggest  some potential strategies for preventing unstable models.

## Equi-probability models

As a baseline for comparisons, consider a simplistic model for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with uniform probabilities over the sample space, say $P_{\boldsymbol \theta_N}(\boldsymbol x)= |\mathcal{X}|^{-N}$, $\boldsymbol x \in \mathcal{X}^N$, where each random variable has $|\mathcal{X}| \geq 1$ outcomes. In contrast to instability, model probabilities here are completely insensitive to changes in data outcomes across the sample space, and the associated log-ratio of extreme probabilities \@ref(eq:elpr) is
$$
\frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\mbox{uniform probability model}),
$$ 
which is as small as possible. In fact, a LREP value of zero can only occur for a FOES model having uniform probabilities, and such equi-probability models are always S-stable.

## One-parameter exponential models {#one-param-exp}
